---
title: "Google Data Analytics Capstone"
author: "Vedic Yadav"
date: "26/06/2021"
output: prettydoc::html_pretty
vignette: > 
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


This is the case study 1 from the Google Data Analytics Certificate(Cyclistic).It was inspired by Kevin Hartman's case study "'Sophisticated, Clear, and Polished': Divvy and Data Visualization" (available here: <https://artscience.blog/home/divvy-dataviz-case-study>). The case study will be based on the Divvy dataset.


Our motto is to combine the downloaded Divvy data into a single dataframe, then perform basic analysis to answer the essential question which is: “How do members and casual riders use Divvy bikes differently?”

***

> # PACKAGES REQUIRED
The packages required for the analysis are as follows: - <br>
1. tidyverse for data import and wrangling. <br>
2. libridate for date functions. <br>
3. ggplot for visualization. <br>

```{r}
library(tidyverse)  
library(lubridate)  
library(ggplot2)  
getwd() #Shows the current directory
```

> # STEP 1: COLLECTION OF DATA.


* After importing the dataset into the R environment. We want to transform the .csv file for each quarter of the year into a single dataframe.
```{r}
q2_2019 <- read_csv("Divvy_Trips_2019_Q2.csv")
q3_2019 <- read_csv("Divvy_Trips_2019_Q3.csv")
q4_2019 <- read_csv("Divvy_Trips_2019_Q4.csv")
q1_2020 <- read_csv("Divvy_Trips_2020_Q1.csv")
````

> # STEP 2: COMBINING THE DATA INTO A SINGLE DATAFRAME.


* Inspecting the columns in each of the dataframe.
```{r}
colnames(q3_2019)
colnames(q4_2019)
colnames(q2_2019)
colnames(q1_2020)
```

* After inspecting the dataframes, we can see that the number of columns in the dataframe for Q1_2020 is more than other dataframes.

* Rename columns to make them consistent with q1_2020
```{r}
(q4_2019 <- rename(q4_2019
                   ,ride_id = trip_id
                   ,rideable_type = bikeid 
                   ,started_at = start_time  
                   ,ended_at = end_time  
                   ,start_station_name = from_station_name 
                   ,start_station_id = from_station_id 
                   ,end_station_name = to_station_name 
                   ,end_station_id = to_station_id 
                   ,member_casual = usertype))

(q3_2019 <- rename(q3_2019
                   ,ride_id = trip_id
                   ,rideable_type = bikeid 
                   ,started_at = start_time  
                   ,ended_at = end_time  
                   ,start_station_name = from_station_name 
                   ,start_station_id = from_station_id 
                   ,end_station_name = to_station_name 
                   ,end_station_id = to_station_id 
                   ,member_casual = usertype))

(q2_2019 <- rename(q2_2019
                   ,ride_id = "01 - Rental Details Rental ID"
                   ,rideable_type = "01 - Rental Details Bike ID" 
                   ,started_at = "01 - Rental Details Local Start Time"  
                   ,ended_at = "01 - Rental Details Local End Time"  
                   ,start_station_name = "03 - Rental Start Station Name" 
                   ,start_station_id = "03 - Rental Start Station ID"
                   ,end_station_name = "02 - Rental End Station Name" 
                   ,end_station_id = "02 - Rental End Station ID"
                   ,member_casual = "User Type"))

```

* Converting "ride_id" and "rideable_type" to character so that they can stack correctly.
```{r}
q4_2019 <-  mutate(q4_2019, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type)) 
q3_2019 <-  mutate(q3_2019, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type)) 
q2_2019 <-  mutate(q2_2019, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type)) 

```

* Now, we will stack individual quarter's data frames into one bigger data frame.
```{r}
all_trips <- bind_rows(q2_2019, q3_2019, q4_2019, q1_2020)
```

* Remove columns named lat, long, birthyear, and gender fields as this data was dropped beginning in 2020.
```{r}
all_trips <- all_trips %>%  
  select(-c(start_lat, start_lng, end_lat, end_lng, birthyear, gender, "01 - Rental Details Duration In Seconds Uncapped", "05 - Member Details Member Birthday Year", "Member Gender", "tripduration"))
```

> # STEP 3: CLEAN UP DATA AND PREPARE FOR ANALYSIS.

* Inspecting the new table that has been created.

```{r}
colnames(all_trips)  #List of column names
nrow(all_trips)  #How many rows are in the data frame?
dim(all_trips)  #Dimensions of the data frame?
head(all_trips)  #See the first 6 rows of data frame.  Also tail(qs_raw)
str(all_trips)  #See list of columns and data types (numeric, character, etc)
summary(all_trips)  #Statistical summary of data. Mainly for numerics
```

* After inspecting the dateset we can see that there are some problems to be fixed:
  + In the "member_casual" column, there are two names for members ("member" and "Subscriber") and two names for casual riders ("Customer" and "casual"). We will need to merge that from four to two columns. <br> 
  + The data can only be consolidated at the ride-level, which is too granular. We will want to add some additional columns of data -- such as day, month, year -- that provide additional opportunities to aggregate the data. <br> 
  + We will want to add a calculated field for length of ride since the 2020Q1 data did not have the "tripduration" column. We will add "ride_length" to the entire dataframe for consistency. <br> 
  + There are some rides where "tripduration" shows up as negative, including several hundred rides where Divvy took bikes out of circulation for Quality Control reasons. We will want to delete these rides. <br>


* In the "member_casual" column, replacing "Subscriber" with "member" and "Customer" with "casual"

* Begin by seeing how many observations fall under each usertype
```{r}
table(all_trips$member_casual)
```

* Now, will reassign the column names as per the 2020_Q1 format.
```{r}
all_trips <-  all_trips %>% 
  mutate(member_casual = recode(member_casual
                           ,"Subscriber" = "member"
                           ,"Customer" = "casual"))

table(all_trips$member_casual)
```

* Add columns that for date, month, day, and year of each ride.
* This will allow us to aggregate ride data for each month, day, or year ... before completing these operations we could only aggregate at the ride level

```{r}
all_trips$date <- as.Date(all_trips$started_at) #The default format is yyyy-mm-dd
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")
```

* Adding a column named "ride_length" for the calculation of all_trips (in seconds)
```{r}
all_trips$ride_length <- difftime(all_trips$ended_at,all_trips$started_at)
str(all_trips)
```
* Convert the column "ride_length" from Factor to numeric so we can run calculations on the dataframe.

```{r}
is.factor(all_trips$ride_length)
all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))
is.numeric(all_trips$ride_length)
```

* On inspecting we found that some of the bikes were taken out for the inspection, hence having ride_length < 0, thus we will have to remove those values from the dateframe(calling it v2).

```{r}
all_trips_v2 <- all_trips[!(all_trips$start_station_name == "HQ QR" | all_trips$ride_length<0),]
```

> # STEP 4: CONDUCTING DESCRIPTIVE ANALYSIS

*Descriptive analysis on ride_length (all figures in seconds)

```{r}
mean(all_trips_v2$ride_length) #average time of ride(total ride length / rides)
median(all_trips_v2$ride_length) #midpoint number in the ascending array of ride lengths
max(all_trips_v2$ride_length) #longest ride in the year
min(all_trips_v2$ride_length) #shortest ride in the year
```


* Compare members and casual users
```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = mean)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = median)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = max)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = min)
```

* See the average ride time by each day for members vs casual users
```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
```

* Notice that the days of the week are out of order. Let's fix that.
```{r}
all_trips_v2$day_of_week <- ordered(all_trips_v2$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
```

* Now, let's run the average ride time by each day for members vs casual users
```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
```

* Analyze ridership data by type and weekday
```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>%  #creates weekday field using wday()
  group_by(member_casual, weekday) %>%  #groups by usertype and weekday
  summarise(number_of_rides = n()							#calculates the number of rides and average duration 
  ,average_duration = mean(ride_length)) %>% 		# calculates the average duration
  arrange(member_casual, weekday)								# sorts
```

* Let's visualize the number of rides by rider type
```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge")
```

* Let's create a visualization for average duration
```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge")
```



